---
title: "Introduction to R timeseries"
author: "Quan Nguyen"
date: "May 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We will use the Canadian car sales data to study a time series.

The data has been extracted from:

http://www.goodcarbadcar.net/2012/10/canada-overall-auto-industry-sales-figures.html

It contains monthly sales data of number of cars sold in Canada since January 2010.

## Prepare the data

```{r}
carsales = read.csv("canadian_carsales.csv", header=T, stringsAsFactors = F)
str(carsales)
# convert Sales column to numeric
carsales$Sales = as.numeric(gsub(",", "", carsales$Sales))
# the x-axis does not show the year/month
plot(carsales$Sales, type='l')
```

## Convert data frame into a timeseries using ts() function

```{r}
carsales.ts = ts(carsales$Sales, frequency=12, start=c(2010,1))
str(carsales.ts) 
# the x-axis should now show the year/month
plot(carsales.ts)
```

## Time series decomposition using stl() function

```{r}
carsales.stl = stl(carsales.ts, s.window = 4)
plot(carsales.stl)
```

## Arima forecast of the ts timeseries

```{r}
library(forecast)
carsales.forecast = forecast.Arima(auto.arima(carsales.ts))

# Compare actual, auto.arima and forecast
plot.forecast(carsales.forecast, col="blue", xlab="Year", sub="Canadian Car Sales")
lines(carsales.forecast$fitted, col="red", lty=2)
lines(carsales.ts, col="black")
legend(
  'topleft', inset=.02, 
  legend=c("Forecasted", "auto.Arima-Fitted", "Original"),
  col=c("blue", "red", "black"),
  lty=1, box.lty=0, cex=0.8
)
# another way to plot the forevast using ggplot2's autoplot()
library(ggplot2)
autoplot(carsales.forecast) 
```

## Create some chaos, delete sample # 32 and 55

```{r}
carsales_partial = carsales[-c(32, 55),]
# We won't be able to tell the missing samples from a non-timeseries plot
plot(carsales_partial$Sales, type='h')
```

## Create a timeseries using zoo() as ts() cannot be used to create a timeseries with missing samples

```{r}
library(zoo)
# create a date column to be used as time index
carsales_partial$date = as.Date(
  paste0(carsales_partial$Year,'-', trimws(carsales_partial$Month), '-01'),
  format="%Y-%B-%d"
)
carsales_partial.z = zoo(
  carsales_partial$Sales, 
  order.by=carsales_partial$date,
  frequency=12
)
str(carsales_partial.z)
# view the index portion
head(index(carsales_partial.z))
# view the data value portion
head(coredata(carsales_partial.z))
plot(carsales_partial.z, type="h")
# get date range from the timeseries
sub_text = range(index(carsales_partial.z))
```

``` {r fig.height=12, fig.width=12}
# compare with good timeseries
par(mfrow=c(2,1))
plot(carsales.ts, col="red", type='h')
plot(carsales_partial.z, col="blue", type='h')
par(mfrow=c(1,1))
```

## Fix the missing samples by imputing

Replace the missing values with the na.locf() function (last observation carried forward)

**Fixing is required for time series decomposition and Arima forecast.**

```{r}
# generate a sequence of monthly intervals
g = seq(start(carsales_partial.z), end(carsales_partial.z), by="month")
carsales_fixed.z = na.locf(carsales_partial.z, xout=g)
```

``` {r fig.height=12, fig.width=12}
# compare the missing sample series with the fixed series
par(mfrow=c(2,1))
plot(carsales_partial.z, col="blue", type='h')
plot(carsales_fixed.z, col="green", type='h')
par(mfrow=c(1,1))
carsales_fixed.z
```

## Timeseries decomposition with decompose() of the fixed zoo object

We will try additive and then multiplicative decompose

```{r}
carsales_fixed.decompose = decompose(
  ts(
    carsales_fixed.z, 
    frequency = 12,
    start=c(2010,1) # we cheat by hard coding the value here
  ),
  "additive"
)
#
plot(carsales_fixed.decompose, xlab="Year")
title(sub=paste(sub_text,collapse="  to  "))

carsales_fixed.decompose = decompose(
  ts(
    carsales_fixed.z, 
    frequency = 12,
    start=c(2010,1) # we cheat by hard coding the value here
  ),
  "multiplicative"
)
#
plot(carsales_fixed.decompose, xlab="Year")
title(sub=paste(sub_text,collapse="  to  "))
```

## Arima forecast

```{r}
carsales_fixed.z.forecast = forecast.Arima(
  auto.arima(  
    ts(
      carsales_fixed.z, 
      frequency = 12,
      start=c(2010,1) # we cheat by hard coding the value here
    )
  )
)
# Compare actual, auto.arima and forecast
plot.forecast(carsales_fixed.z.forecast, col="blue")
lines(carsales_fixed.z.forecast$fitted, col="red", lty=2)
lines(carsales.ts, col="black")
legend(
  'topleft', inset=.02, 
  legend=c("Forecasted", "auto.Arima-Fitted", "Original"),
  col=c("blue", "red", "black"),
  lty=1, box.lty=0, cex=0.8
)
# Another way to plot arima forecast with autoplot()
library(ggplot2)
autoplot(carsales_fixed.z.forecast) 
```

